---
title: "K-means clustering"
author: "Simone"
date: "March 14, 2020"
output: ioslides_presentation
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 5)
```

## Purpose of this App

This App allows users to generate random datasets composed of up to 5 clusters, each one normally distributed across 3 dimensions.  
  
Additionally, it is possible to run a 'K-means clustering' algortihm on the datasets.  
  
The App can be tested here:  
  
- [https://simonedut.shinyapps.io/K-means-clustering](https://simonedut.shinyapps.io/K-means-clustering)
  
While all the code is available here:  
  
- [https://github.com/SimoneDut/K-means-clustering](https://github.com/SimoneDut/K-means-clustering)

## Dataset generation (1)

A dataset will be generated at first, by setting a number of parameters.  
  
Each point in the dataset will be part of a cluster in a 3 dimensional space, and each dimension of the cluster will be normally distributed using the following R expression:

```{r chunk_1, echo = TRUE, eval = FALSE}
cluster_dimension <- rnorm(cluster_elements,
                           mean = cluster_mean,
                           sd = cluster_standard_deviation)
```

## Dataset generation (2)

Once the dataset is generated, it will be displayed as follows:

```{r chunk_2}
library(shiny)
library(plotly)
library(RColorBrewer)

updated <- list()
data <- list()
for (i in 1:3) {
  for (l in c("x", "y", "z")) {
    updated[[paste(l, i, sep = "")]] <- c(30, 50, 70)[i]
    updated[[paste("s", l, i, sep = "")]] <- 10
  }
}
updated$c <- 3
updated$m <- 100
data_app <- data.frame()
for (i in 1:updated$c) {
  temp <- list()
  for (l in c("x", "y", "z")) {
    temp[[l]] <- rnorm(updated$m,
                       mean = updated[[paste(l, i, sep = "")]],
                       sd = updated[[paste("s", l, i, sep = "")]])
  }
  temp[["c"]] <- rep(i, updated$m)
  temp[["k"]] <- rep(0, updated$m)
  data_app <- rbind(data_app, data.frame(temp))
}
data$x <- data_app$x
data$y <- data_app$y
data$z <- data_app$z
data$c <- data_app$c
data$s <- ifelse(data_app$c == 5, 10, 12)
data$k <- data_app$k
```

```{r chunk_3}
colors <- c("black", brewer.pal(9, "Set1"), "lightblue")
n_colors <- dim(table(data$k))
if (min(data$k) == 0) {
  colors <- colors[1:n_colors]
} else {
  colors <- colors[2:(n_colors+1)]
}
div(plot_ly(x = data$x, y = data$y, z = data$z,
            type = "scatter3d",
            mode = "markers",
            color = as.factor(data$k),
            colors = colors,
            symbol = as.factor(data$c),
            symbols = c("circle", "cross", "square", "diamond", "x"),
            size = data$s,
            width = 400, height = 400) %>%
      layout(scene = list(xaxis = list(title = "Average transaction value", range = c(0, 100)),
                          yaxis = list(title = "Transactions per year", range = c(0, 100)),
                          zaxis = list(title = "Customer seniority", range = c(0, 100)),
                          camera = list(eye = list(x = 1.6,
                                                   y = 1.6,
                                                   z = 0.5))),
             showlegend = FALSE),
    align = "center")
```

## Results after running the algorithm

Clusters will be marked using different colors:

```{r chunk_4}
data_app <- data.frame(x = data$x, y = data$y, z = data$z, k = data$k)
random_init_indices <- sample(1:nrow(data_app), 3)
centroids <- list()
for (i in 1:3) {
  centroids[[i]] <- data.frame(x = data_app$x[random_init_indices[i]],
                               y = data_app$y[random_init_indices[i]],
                               z = data_app$z[random_init_indices[i]])
}
distances_app <- list()
for (i in 1:20) {
  ## Calculate the nearest centroid for each point (Assignment step)
  for (j in 1:3) {
    distances_app[[paste("dist", j, sep = "_")]] <- sqrt((data_app$x - centroids[[j]]$x[1])^2 +
                                                         (data_app$y - centroids[[j]]$y[1])^2 +
                                                         (data_app$z - centroids[[j]]$z[1])^2)
  }
  if (class(distances_app) == "list") { distances_app <- data.frame(distances_app) }
  data_app$k <- apply(distances_app, 1, which.min)
  ## Recalculate the position of the centroids (Update step)
  lost_centroids <- integer(0)
  for (j in 1:3) {
    nearest_points <- data_app[data_app$k == j,]
    n_points <- dim(nearest_points)[1]
    if (n_points > 0) {
      centroids[[j]] <- rbind(data.frame(x = sum(nearest_points$x) / n_points,
                                         y = sum(nearest_points$y) / n_points,
                                         z = sum(nearest_points$z) / n_points),
                              centroids[[j]])
    } else { lost_centroids <- c(lost_centroids, j) }
    }
  ## Reinitialize lost centroids
  random_init_indices <- sample(1:nrow(data_app), length(lost_centroids))
  for (j in lost_centroids) {
    centroids[[j]] <- rbind(data.frame(x = data_app$x[random_init_indices[j]],
                                       y = data_app$y[random_init_indices[j]],
                                       z = data_app$z[random_init_indices[j]]),
                            centroids[[j]])
  }
}
data$k <- data_app$k
```

```{r chunk_5, fig.width = 2, fig.height = 2, fig.align = "center"}
colors <- c("black", brewer.pal(9, "Set1"), "lightblue")
n_colors <- dim(table(data$k))
if (min(data$k) == 0) {
  colors <- colors[1:n_colors]
} else {
  colors <- colors[2:(n_colors+1)]
}
div(plot_ly(x = data$x, y = data$y, z = data$z,
            type = "scatter3d",
            mode = "markers",
            color = as.factor(data$k),
            colors = colors,
            symbol = as.factor(data$c),
            symbols = c("circle", "cross", "square", "diamond", "x"),
            size = data$s,
            width = 400, height = 400) %>%
      layout(scene = list(xaxis = list(title = "Average transaction value", range = c(0, 100)),
                          yaxis = list(title = "Transactions per year", range = c(0, 100)),
                          zaxis = list(title = "Customer seniority", range = c(0, 100)),
                          camera = list(eye = list(x = 1.6,
                                                   y = 1.6,
                                                   z = 0.5))),
             showlegend = FALSE),
    align = "center")
```
